---
title: "Chapter 4: Doing data analysis with the multilevel model for change"
---

::: {.alert .alert-warning}
This chapter is under construction.
:::

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(
  panel.grid.major = ggplot2::element_blank(),
  panel.grid.minor = ggplot2::element_blank()
)
```

```{r setup}
library(alda)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(patchwork)
library(lme4)
library(broom.mixed)
```

## 4.1 Example: Changes in adolescent alcohol use

Figure 4.1, page 77:

```{r}
alcohol_use_1 |>
  filter(id %in% c(4, 14, 23, 32, 41, 56, 65, 82)) |>
  ggplot(aes(x = age, y = alcohol_use)) +
    stat_smooth(method = "lm", se = FALSE) +
    geom_point() +
    coord_cartesian(xlim = c(13, 17), ylim = c(0, 4)) +
    facet_wrap(vars(id), ncol = 4)
```

Figure 4.2, page 79:

> To construct this display, we twice divided this subsample into two groups: once by COA (top panel) and again by PEER (bottom panel). Because PEER is continuous, the bottom panel represents a split at the sample mean. Thicker lines represent coincident trajectories—the thicker the line, the more trajectories.

```{r}
figure_4.2 <- map(
  set_names(c("child_of_alcoholic", "peer_alcohol_use")),
  \(.x) {
    # Wrangle
    subset_data <- alcohol_use_1 |>
      # In order to count the number of coincident trajectories, the data needs
      # to be in wide format so we can group by each time point.
      pivot_wider(names_from = age, values_from = alcohol_use) |>
      mutate(
        peer_alcohol_use = if_else(peer_alcohol_use < 1.01756, "low", "high"),
        peer_alcohol_use = factor(peer_alcohol_use, levels = c("low", "high"))
      ) |>
      # Indexing into the .data pronoun allows us to use group_by() with
      # the character string inputs ("child_of_alcoholic", "peer_alcohol_use").
      group_by(.data[[.x]], `14`, `15`, `16`) |>
      summarise(n = n()) |>
      ungroup() |>
      # An ID is needed to indicate which observations go together after pivoting
      # back to long format.
      mutate(id = 1:n()) |>
      pivot_longer(
        cols = starts_with("1"),
        names_to = "age",
        names_transform = as.integer,
        values_to = "alcohol_use"
      )
    
    # Plot
    ggplot(subset_data, aes(x = age, y = alcohol_use, group = id)) +
      stat_smooth(
        aes(linewidth = n), method = "lm", se = FALSE, show.legend = FALSE
      ) +
      scale_linewidth_continuous(range = c(.25, 4)) +
      coord_cartesian(xlim = c(13, 17), ylim = c(0, 4)) +
      # To learn about the .data pronoun, see:
      # https://dplyr.tidyverse.org/articles/programming.html
      facet_wrap(vars(.data[[.x]]), labeller = label_both) +
      # We're going to create a shared axis label with patchwork later, see:
      # https://tidytales.ca/snippets/2022-12-22_patchwork-shared-axis-labels/
      ylab(NULL)
  }
)

# Before plotting, remove redundant axis elements for aesthetics and set margins
# so it looks exactly the same as if the plot was created with facet_wrap(). The
# margin values selected here are based on the default theme values.
patch <- wrap_plots(figure_4.2, nrow = 2) &
  theme(plot.margin = margin(0, 0, 0, 0))
patch[[1]] <- patch[[1]] +
  xlab(NULL) +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    plot.margin = margin(5.5, 5.5, 1.375, 0)
  )
patch[[2]] <- patch[[2]] +
  theme(plot.margin = margin(1.375, 5.5, 5.5, 0))

# Create a pseudo y-axis title using a tag.
wrap_elements(patch) +
  labs(tag = "alcohol_use") +
  theme(
    plot.tag = element_text(
      size = rel(1), angle = 90, vjust = 1,
      margin = margin(b = 8.25, l = 5.5, r = 2.75)
    ),
    plot.tag.position = "left",
    plot.margin = margin(0)
  )
```

## 4.2 The composite specification of the multilevel model for change

## 4.3 Methods of Estimation, Revisited

## 4.4 First Steps: Fitting Two Unconditional Multilevel Models for Change

Table 4.1, page 94-95:

```{r}
# Fit models:
model_A <- lmer(
  alcohol_use ~ 1 + (1 | id),
  data = alcohol_use_1,
  REML = FALSE
)

model_B <- update(
  model_A,
  . ~ . - (1 | id) + I(age - 14) + (I(age - 14) | id)
)

model_C <- update(
  model_B,
  . ~ . + child_of_alcoholic + I(age - 14):child_of_alcoholic
)

model_D <- update(
  model_C,
  . ~ . + peer_alcohol_use + I(age - 14):peer_alcohol_use
)

model_E <- update(
  model_D,
  . ~ . - I(age - 14):child_of_alcoholic
)

model_F <- update(
  model_B,
  . ~ . + child_of_alcoholic +
    I(peer_alcohol_use - 1.018) + I(age - 14):I(peer_alcohol_use - 1.018)
)

model_G <- update(
  model_F,
  . ~ . - child_of_alcoholic + I(child_of_alcoholic - 0.451)
)

# Collect models together for purrr:
table_4.1_models <- set_names(
  list(model_A, model_B, model_C, model_D, model_E, model_F, model_G),
  LETTERS[1:7]
)

# Estimates:
map_dfr(table_4.1_models, tidy, .id = "model") |>
  select(-statistic) |>
  pivot_longer(cols = c(estimate, std.error), names_to = "statistic") |>
  pivot_wider(names_from = model, values_from = value) |>
  arrange(effect, desc(group)) |>
  print(n = 26)

# Goodness of fit:
map_dfr(table_4.1_models, glance, .id = "model")
```

Figure 4.3, page 99:

```{r}
# Model B
prototypical_alcohol_users_B <- tibble(age = 14:16)

prototypical_alcohol_use_B <- tibble(
  alcohol_use = predict(
    model_B,
    prototypical_alcohol_users_B,
    re.form = NA
  )
)

modelbased::estimate_prediction(model_B, data = tibble(age = 14:16)) |>
  ggplot(aes(x = age, y = Predicted)) +
    geom_line() +
    scale_x_continuous(breaks = 13:17) +
    coord_cartesian(xlim = c(13, 17), ylim = c(0, 2)) +
    labs(y = "alcohol_use")

prototypical_alcohol_users_B |>
  bind_cols(prototypical_alcohol_use_B) |>
  ggplot(aes(x = age, y = alcohol_use)) +
    geom_line() +
    scale_x_continuous(breaks = 13:17) +
    coord_cartesian(xlim = c(13, 17), ylim = c(0, 2))

# Model C
prototypical_alcohol_users_C <- crossing(
  age = 14:16,
  child_of_alcoholic = 0:1
)

prototypical_alcohol_use_C <- tibble(
  alcohol_use = predict(
    model_C,
    prototypical_alcohol_users_C,
    re.form = NA
  )
)

prototypical_alcohol_users_C |>
  bind_cols(prototypical_alcohol_use_C) |>
  mutate(child_of_alcoholic = factor(child_of_alcoholic)) |>
  ggplot(aes(x = age, y = alcohol_use, colour = child_of_alcoholic)) +
    geom_line() +
    scale_x_continuous(breaks = 13:17) +
    coord_cartesian(xlim = c(13, 17), ylim = c(0, 2))

modelbased::estimate_prediction(
  model_C, data = crossing(age = 14:16, child_of_alcoholic = 0:1)
) |>
  rename(alcohol_use = Predicted) |>
  mutate(child_of_alcoholic = factor(child_of_alcoholic)) |>
  ggplot(aes(x = age, y = alcohol_use, colour = child_of_alcoholic)) +
    geom_line() +
    scale_x_continuous(breaks = 13:17) +
    coord_cartesian(xlim = c(13, 17), ylim = c(0, 2))

# Model E
prototypical_alcohol_users_E <- crossing(
  age = 14:16,
  child_of_alcoholic = 0:1,
  peer_alcohol_use = c(0.655, 1.381)
)

prototypical_alcohol_use_E <- tibble(
  alcohol_use = predict(
    model_E,
    prototypical_alcohol_users_E,
    re.form = NA
  )
)

prototypical_alcohol_users_E |>
  bind_cols(prototypical_alcohol_use_E) |>
  mutate(child_of_alcoholic = factor(child_of_alcoholic)) |>
  ggplot(aes(x = age, y = alcohol_use, colour = child_of_alcoholic)) +
    geom_line() +
    scale_x_continuous(breaks = 13:17) +
    coord_cartesian(xlim = c(13, 17), ylim = c(0, 2)) +
    facet_wrap(vars(peer_alcohol_use), labeller = label_both)
```

## 4.5 Practical Data Analytic Strategies for Model Building

## 4.6 Comparing Models Using Deviance Statistics

## 4.7 Using Wald Statistics to Test Composite Hypotheses About Fixed Effects

## 4.8 Evaluating the Tenability of a Model’s Assumptions

Figure 4.4:

```{r}
ols_fit <- lmList(
  alcohol_use ~ I(age - 14) | id,
  pool = FALSE,
  data = alcohol_use_1
)

ols_fit |>
  map_dfr(tidy, .id = "id") |>
  select(-c(std.error, statistic, p.value)) |>
  #pivot_longer(cols = c(estimate, names_to = "statistic") |>
  left_join(
    alcohol_use_1 |>
      select(id, child_of_alcoholic, peer_alcohol_use) |>
      distinct() |>
      mutate(id = as.character(id))
  ) |>
  pivot_longer(
    cols = c(child_of_alcoholic, peer_alcohol_use),
    names_to = "predictor"
  ) |>
  ggplot(aes(x = value, y = estimate)) +
    geom_hline(yintercept = 0, alpha = .25) +
    geom_point(alpha = .5) +
    facet_wrap(vars(predictor, term), dir = "v", scales = "free") +
    ggh4x::facetted_pos_scales(
      x = list(
        predictor == "child_of_alcoholic" ~
          scale_x_continuous(breaks = c(0, 1), limits = c(-1, 2)),
        predictor == "peer_alcohol_use" ~
          scale_x_continuous(limits = c(0, 3))
      ),
      y = list(
        term == "I(age - 14)" ~ scale_y_continuous(limits = c(-1, 2))
      )
    )
```

Figure 4.5, page 131:

```{r}
# Left side figures
library(easystats)
performance::check_normality(model_F, effects = "fixed") |> plot(type = "qq")
performance::check_normality(model_F, effects = "random") |> plot(type = "qq")
```

Figure 4.6, page 133:

```{r}
# Fixed effects
model_F |>
  augment(re.form = NA) |>
  ggplot(aes(x = id, y = scale(.resid))) +
  geom_point() +
  coord_cartesian(ylim = c(-3, 3))

# Random effects
model_F |>
  ranef() |>
  as_tibble() |>
  rename(id = grp) |>
  left_join(
    alcohol_use_1 |>
      select(id, child_of_alcoholic, peer_alcohol_use) |>
      distinct() |>
      mutate(id = factor(id))
  ) |>
  pivot_longer(
    cols = c(child_of_alcoholic, peer_alcohol_use), names_to = "predictor"
  ) |>
  ggplot(aes(x = value, y = condval)) +
    geom_point() +
    scale_y_continuous(limits = c(-1, 1)) +
    facet_wrap(vars(term, predictor), scales = "free_x") +
    ggh4x::facetted_pos_scales(
      x = list(
        predictor == "child_of_alcoholic" ~
          scale_x_continuous(breaks = c(0, 1), limits = c(-1, 2)),
        predictor == "peer_alcohol_use" ~
          scale_x_continuous(limits = c(0, 3))
      )
    )
```

## 4.9 Model-Based (Empirical Bayes) Estimates of the Individual Growth Parameters

Figure 4.7:

```{r}
alcohol_use_1_subset <- alcohol_use_1 |>
  filter(id %in% c(4, 14, 23, 32, 41, 56, 65, 82))

alcohol_use_1_subset |>
  mutate(
    .fixed  = predict(model_F, newdata = alcohol_use_1_subset, re.form = NA),
    .fitted = predict(model_F, newdata = alcohol_use_1_subset)
  ) |>
  ggplot(aes(x = age)) +
    geom_point(aes(y = alcohol_use)) +
    # The OLS lines can just be fit here, instead of going through the trouble
    # of fitting them separately.
    stat_smooth(
      aes(y = alcohol_use),
      method = "lm", se = FALSE, linetype = 2, linewidth = .5
    ) +
    geom_line(
      aes(y = .fixed), linewidth = .5
    ) +
    geom_line(
      aes(y = .fitted), linewidth = 1
    ) +
    scale_y_continuous(breaks = 0:4, limits = c(-1, 4)) +
    xlim(13, 17) +
    facet_wrap(vars(id), nrow = 2, labeller = label_both)
```
