---
title: "Chapter 6: Modeling Discontinuous and Nonlinear Change"
---

::: {.alert .alert-warning}
This chapter is under construction.
:::

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(
  panel.grid.major = ggplot2::element_blank(),
  panel.grid.minor = ggplot2::element_blank()
)
```

```{r setup}
library(alda)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)
library(lme4)
```

## 6.1 Discontinuous Individual Change

Table 6.1, page 192:

```{r}
dropout_wages |>
  filter(id %in% c(206,2365,4384)) |>
  select(id, log_wages, experience, ged, postsecondary_education) |>
  mutate(ged_x_experience = ged * experience)
```

Table 6.2, page 203:

```{r}
wages_fit_A <- lmer(
  log_wages ~
    experience +
    I(highest_grade - 9) +
    experience:black +
    I(unemployment_rate - 7) +
    (1 + experience | id),
  data = dropout_wages,
  REML = FALSE
)

deviance(wages_fit_A)

wages_fit_B <- update(
  wages_fit_A,
  . ~ . - (1 + experience | id) + ged + (1 + experience + ged | id),
  control = lmerControl(optimizer = "bobyqa")
)

deviance(wages_fit_B)

wages_fit_C <- update(
  wages_fit_A,
  . ~ . + ged,
  control = lmerControl(optimizer = "bobyqa")
)

deviance(wages_fit_C)

wages_fit_D <- update(
  wages_fit_A,
  . ~ 
    . - 
    (1 + experience | id) + 
    postsecondary_education + 
    (1 + experience + postsecondary_education | id),
  control = lmerControl(optimizer = "bobyqa")
)

deviance(wages_fit_D)

wages_fit_E <- update(
  wages_fit_A,
  . ~ . + postsecondary_education
)

deviance(wages_fit_E)

wages_fit_F <- update(
  wages_fit_A,
  . ~ 
    . -
    (1 + experience | id) + 
    ged + 
    postsecondary_education + 
    (1 + experience + ged + postsecondary_education | id),
  control = lmerControl(optimizer = "bobyqa")
)

deviance(wages_fit_F)

wages_fit_G <- update(
  wages_fit_F,
  . ~
    . - 
    (1 + experience + ged + postsecondary_education | id) +
    (1 + experience + ged | id)
)

deviance(wages_fit_G)

wages_fit_H <- update(
  wages_fit_F,
  . ~
    . - 
    (1 + experience + ged + postsecondary_education | id) +
    (1 + experience + postsecondary_education | id)
)

deviance(wages_fit_H)

# FIXME: Singular
wages_fit_I <- update(
  wages_fit_A,
  . ~ 
    . -
    (1 + experience | id) +
    ged +
    experience:ged +
    (1 + experience + ged + experience:ged | id)
)

deviance(wages_fit_I)

wages_fit_J <- update(
  wages_fit_I,
  . ~ 
    . -
    (1 + experience + ged + experience:ged | id) +
    (1 + experience + ged | id)
)

deviance(wages_fit_J)

anova(wages_fit_A, wages_fit_B, wages_fit_D)
anova(wages_fit_B, wages_fit_C, wages_fit_F, wages_fit_I)
anova(wages_fit_D, wages_fit_E, wages_fit_F)
anova(wages_fit_F, wages_fit_G, wages_fit_H)
anova(wages_fit_I, wages_fit_J)
```

Table 6.3, page 205:

```{r}
summary(wages_fit_F)
```

Figure 6.3:

```{r}
prototypical_dropouts <- tibble(
  experience = rep(c(0, 3, 3, 12), times = 4),
  highest_grade = rep(c(9, 12), each = 4, times = 2),
  black = rep(c(FALSE, TRUE), each = 8),
  ged = rep(c(0, 0, 1, 1), times = 4),
  unemployment_rate = 7
) |>
  mutate(postsecondary_education = ifelse(ged == 0, 0, experience - 3))

prototypical_dropout_scores <- tibble(
  log_wages = predict(
    wages_fit_F,
    prototypical_dropouts,
    re.form = NA
  )
)

prototypical_dropouts |>
  bind_cols(prototypical_dropout_scores) |>
  ggplot(aes(x = experience, y = log_wages, colour = as.logical(black))) +
    geom_line() +
    scale_x_continuous(breaks = seq(0, 10, by = 2)) +
    coord_cartesian(xlim = c(0, 10), ylim = c(1.6, 2.4)) +
    facet_wrap(vars(highest_grade))
```

## 6.2 Using Transformations to Model Nonlinear Individual Change

Figure 6.4, page 209:

```{r}
alcohol_fit <- lmer(
  alcohol_use ~ 
    I(age - 14) * peer_alcohol_use + child_of_alcoholic + (1 + I(age - 14) | id),
  data = alcohol_use_1,
  REML = FALSE
)

prototypical_alcohol_users <- tibble(
  age = rep(seq(14, 16, by = .25), times = 4),
  child_of_alcoholic = rep(c(0, 1), each = 9, times = 2),
  peer_alcohol_use = rep(c(.655, 1.381), each = 18)
)

prototypical_alcohol_use <- tibble(
  alcohol_use = predict(
    alcohol_fit,
    prototypical_alcohol_users,
    re.form = NA
  )
)

prototypical_alcohol_users |>
  bind_cols(prototypical_alcohol_use) |>
  mutate(alcohol_use = alcohol_use^2) |>
  ggplot(aes(x = age, y = alcohol_use, colour = factor(peer_alcohol_use))) +
    geom_line() +
    facet_wrap(vars(child_of_alcoholic)) +
    coord_cartesian(xlim = c(13, 17), ylim = c(0, 3))

```

Figure 6.6, page 212:

```{r}
 ggplot(berkeley, aes(x = age, y = iq)) +
  geom_point(data = \(.x) mutate(.x, group = "1: untransformed")) +
  geom_point(data = \(.x) mutate(.x, iq = iq^(2.3), group = "2: iq^(2.3)")) +
  geom_point(data = \(.x) mutate(.x, age = age^(1/2.3), group = "3: age^(1/2.3)")) +
  facet_wrap(vars(group), scales = "free")
```

## 6.3 Representing Individual Change Using a Polynomial Function of TIME

Figure 6.7, page 218:

```{r}
# Polynomial formulas for plotting:
linear <- formula(y ~ x)
quadratic <- formula(y ~ x + I(x^2))
cubic <- formula(y ~ x + I(x^2) + I(x^3))
quartic <- formula(y ~ x + I(x^2) + I(x^3) + I(x^4))

# We need to pmap() over geoms to get facets with unique subsets of data and
# smooths since there's no aesthetic mappings for geom_smooth() formulas.
participant_params <- list(
  list(1, 6, 11, 25, 34, 36, 40, 26),
  LETTERS[1:8],
  list(quadratic, quadratic, linear, linear, cubic, quartic, cubic, quartic)
)

ggplot(mapping = aes(x = grade, y = externalizing_behaviour)) +
  {
    pmap(
      participant_params,
      \(.x, .y, .z) {
        # Subset data to a given id and add a case variable for faceting.
        participant_data <- externalizing_behaviour |>
          filter(id == .x) |>
          mutate(participant = .y)
        
        # Rather than adding the geoms together with + they need to be returned
        # as a list of geoms to be added to the plot from within pmap().
        list(
          geom_point(data = participant_data),
          geom_smooth(
            data = participant_data,
            method = "lm", formula = quartic, se = FALSE,
            linewidth = .25, linetype = 2
          ),
          geom_smooth(
            data = participant_data,
            method = "lm", formula = .z, se = FALSE, linewidth = .5
          )
        )
      }
    )
  } +
  scale_x_continuous(breaks = 0:7) +
  coord_cartesian(xlim = c(0, 7)) +
  facet_wrap(vars(participant), nrow = 2)
```

Table 6.5, page 221:

```{r}
behaviour_fit_A <- lmer(
  externalizing_behaviour ~ 1 + (1 | id),
  data = externalizing_behaviour,
  REML = FALSE
)

summary(behaviour_fit_A)

behaviour_fit_B <- lmer(
  externalizing_behaviour ~ time + (1 + time | id),
  data = externalizing_behaviour,
  REML = FALSE
)

summary(behaviour_fit_B)

behaviour_fit_C <- lmer(
  externalizing_behaviour ~
    poly(time, 2, raw = TRUE) + (1 + poly(time, 2, raw = TRUE) | id),
  data = externalizing_behaviour,
  REML = FALSE
)

summary(behaviour_fit_C)

# Singular
behaviour_fit_D <- lmer(
  externalizing_behaviour ~
    poly(time, 3, raw = TRUE) + (1 + poly(time, 3, raw = TRUE) | id),
  data = externalizing_behaviour,
  REML = FALSE
)

summary(behaviour_fit_D)
```

## 6.4 Truly Nonlinear Trajectories

Figure 6.8, page 227:

```{r}
cognitive_growth |>
  filter(id %in% c(1, 4, 6, 7, 8, 11, 12, 15)) |>
  ggplot(aes(x = game, y = nmoves)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 30), ylim = c(0, 25)) +
  facet_wrap(vars(id), ncol = 4, labeller = label_both)
```

Help wanted for Table 6.6, page 231, see: <https://github.com/mccarthy-m-g/alda/issues/3>

<!--
$$
\begin{alignat}{3}
&\text{Level 1:}& \qquad
  Y_{ij} &= 1 + \frac{19}{1 + \Pi_{0i} e^{-(\Pi_{1i} t_{ij})}} + \epsilon_{ij} \\
  
&\text{Level 2:} \\\\
&\qquad\text{(A)}&
  \Pi_{0i} &= \gamma_{00} + \zeta_{0i} \\&&
  \Pi_{1i} &= \gamma_{10} + \zeta_{1i} \\\\

&\qquad\text{(B)}&
  \Pi_{0i} &= \gamma_{00} + \gamma_{01}(\mathrm{READ}_i - \overline{\mathrm{READ}}) +
              \zeta_{0i} \\&&
  \Pi_{1i} &= \gamma_{10} + \gamma_{11}(\mathrm{READ}_i - \overline{\mathrm{READ}}) + 
              \zeta_{1i}
\end{alignat}
$$
-->

<!--
Table 6.6, page 231 with formulas used in website examples models: <https://stats.oarc.ucla.edu/sas/examples/alda/chapter6/applied-longitudinal-data-analysis-modeling-change-and-event-occurrenceby-judith-d-singer-and-john-b-willett-chapter-6-modeling-discontinuous-and-nonlinear-c/>. Note: These seem to be the models that were actually used in the table, but I don't understand how these were attained given the equations in the textbook.

$$
\begin{alignat}{3}
&\text{Model A:}& \qquad
  Y_{ij} &= 1 + \frac{19}{1 + \gamma_{00} * e^{-(\gamma_{10} + u_1) t + u_0}} +
            \epsilon_{ij} \\\\

&\text{Model B:}& \qquad
  Y_{ij} &= 1 + \frac{19}{1 + \gamma_{00} * e^{-((\gamma_{10} + u_1) t +
            \gamma_{01}x + \gamma_{11}xt + u_0)}} + \epsilon_{ij}
\end{alignat}
$$
-->

Figure 6.10, page 232:

```{r}
# TODO after resolving the equations and code for Models A and B.
```
