---
title: "Chapter 12: Extending the discrete-time hazard model"
---

::: {.alert .alert-warning}
This chapter is under construction.
:::

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(
  panel.grid.major = ggplot2::element_blank(),
  panel.grid.minor = ggplot2::element_blank()
)
```

```{r setup}
library(alda)
library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(ggplot2)
```

## 12.1 Alternative Specifications for the "Main Effect of TIME"

Table 12.2, page 413:

```{r}
# Convert to person-period format
tenure_pp <- tenure |>
  reframe(
    year = 1:max(years),
    event = if_else(year == years & censor == 0, 1, 0),
    .by = id
  ) |>
  mutate(
    temp_year = year,
    temp_dummy = 1
  ) |>
  pivot_wider(
    names_from = temp_year,
    names_prefix = "year_",
    values_from = temp_dummy,
    values_fill = 0
  )

# Fit models
tenure_fit_general <- glm(
  event ~ factor(year), family = "binomial", data = tenure_pp
)

tenure_fit_constant <- glm(
  event ~ 1, family = "binomial", data = tenure_pp
)

tenure_fit_linear    <- update(tenure_fit_constant, . ~ year)
tenure_fit_quadratic <- update(tenure_fit_linear, . ~ . + I(year^2))
tenure_fit_cubic     <- update(tenure_fit_quadratic, . ~ . + I(year^3))
tenure_fit_order_4   <- update(tenure_fit_cubic, . ~ . + I(year^4))
tenure_fit_order_5   <- update(tenure_fit_order_4, . ~ . + I(year^5))

# Compare
anova(
  tenure_fit_constant,
  tenure_fit_linear,
  tenure_fit_quadratic,
  tenure_fit_cubic,
  tenure_fit_order_4,
  tenure_fit_order_5
)
```

Figure 12.1, page 414:

```{r}
tenure_fit_trajectories <- map_df(
  list(
    constant = tenure_fit_constant,
    linear = tenure_fit_linear,
    quadratic = tenure_fit_quadratic,
    cubic = tenure_fit_cubic,
    general = tenure_fit_general
  ),
  \(.x) {
    augment(.x, newdata = tibble(year = 1:9)) 
  },
  .id = "model"
)

tenure_fit_trajectories |>
  mutate(
    model = factor(
      model, levels = c("constant", "linear", "quadratic", "cubic", "general")
    ),
    hazard = if_else(
      model %in% c("quadratic", "general"), 1 / (1 + exp(-.fitted)), NA
    ),
    survival = if_else(
      model %in% c("quadratic", "general"), cumprod(1 - hazard), NA
    ),
    .by = model
  ) |>
  rename(logit_hazard = .fitted) |>
  pivot_longer(cols = logit_hazard:survival, names_to = "estimate") |>
  mutate(estimate = factor(
    estimate, levels = c("logit_hazard", "hazard", "survival"))
  ) |>
  ggplot(aes(x = year, y = value, colour = model)) +
    geom_line() +
    scale_color_brewer(type = "qual", palette = "Dark2") +
    scale_x_continuous(breaks = 1:9) +
    facet_wrap(vars(estimate), scales = "free_y", labeller = label_both)
```

## 12.2 Using the Complementary Log-Log Link to Specify a Discrete-Time Hazard Model

Figure 12.2:

```{r}

```

Figure 12.3, page 423:

```{r}
first_sex_pp <- first_sex |>
  rename(grades = grade) |>
  reframe(
    grade = 7:max(grades),
    event = if_else(grade == grades & censor == 0, 1, 0),
    parental_transition,
    parental_antisociality,
    .by = id
  )

# The nested map_() is used here so we can get an ID column for both the
# link function and the subset.
map_dfr(
  list(logit = "logit", cloglog = "cloglog"),
  \(.x) {
    map_dfr(
      list(`0` = 0, `1` = 1),
      \(.y) {
        first_sex_fit <- glm(
          event ~ factor(grade),
          family = binomial(link = .x),
          data = first_sex_pp,
          subset = c(parental_transition == .y)
        )
        
        augment(first_sex_fit, newdata = tibble(grade = 7:12))
      },
      .id = "parental_transition"
    )
  },
  .id = "link"
) |>
  ggplot(
    aes(x = grade, y = .fitted, colour = parental_transition, linetype = link)
  ) +
    geom_line()

```

Table 12.3, page 424:

```{r}
map_dfr(
  list(cloglog = "cloglog", logit = "logit"),
  \(.x) {
    first_sex_fit <- glm(
      event ~ -1 + factor(grade) + parental_transition,
      family = binomial(link = .x),
      data = first_sex_pp
    )
    
    first_sex_fit |>
      tidy() |>
      select(term, estimate) |>
      mutate(
        base_hazard = case_when(
          .x == "logit" & term != "parental_transition" ~
            1 / (1 + exp(-estimate)),
          .x == "cloglog" & term != "parental_transition" ~
            1 - exp(-exp(estimate))
        )
      )
  },
  .id = "link"
) |>
  pivot_wider(names_from = link, values_from = c(estimate, base_hazard))
```

## 12.3 Time-Varying Predictors

Figure 12.4, page 432:

```{r}
first_depression_fit <- glm(
  depressive_episode ~ poly(I(period - 18), 3, raw = TRUE) + parental_divorce,
  family = binomial(link = "logit"),
  data = first_depression_1
)

# When a predictor enters the model as part of a matrix of covariates, such as
# with stats::poly(), it is represented in augment() as a matrix column. A simple
# workaround to get the predictor on its original scale as a vector is to pass
# the original data to augment().
first_depression_predictions <- first_depression_fit |>
  augment(data = first_depression_1) |>
  mutate(hazard = 1 / (1 + exp(-.fitted)))

# Proportions of the risk set at each age who experienced an initial depressive
# episode at that age, as function of their parental divorce status at that age.
first_depression_proportions <- first_depression_1 |>
  group_by(period, parental_divorce) |>
  summarise(
    total = n(),
    event = sum(depressive_episode),
    proportion = event / total,
    proportion = if_else(proportion == 0, NA, proportion),
    logit = log(proportion / (1 - proportion))
  )

# Top plot
ggplot(mapping = aes(x = period, colour = factor(parental_divorce))) +
  geom_line(
    aes(y = hazard), data = first_depression_predictions
  ) +
  geom_point(
    aes(y = proportion), data = first_depression_proportions
  ) +
  scale_x_continuous(breaks = seq(0, 40, by = 5), limits = c(0, 40)) +
  scale_y_continuous(limits = c(0, 0.06))

# Bottom plot
ggplot(mapping = aes(x = period, colour = factor(parental_divorce))) +
  geom_line(
    aes(y = .fitted), data = first_depression_predictions
  ) +
  geom_point(
    aes(y = logit), data = first_depression_proportions
  ) +
  scale_x_continuous(breaks = seq(0, 40, by = 5), limits = c(0, 40)) +
  scale_y_continuous(breaks = seq(-8, -2, by = 1), limits = c(-8, -2))
```

Figure 12.5, page 437:

```{r}
first_depression_fit_2 <- update(first_depression_fit, . ~ . + female)

first_depression_fit_2 |>
  augment(
    newdata = expand_grid(
      period = 4:39, parental_divorce = c(0, 1), female = c(0, 1)
    )
  ) |>
  mutate(
    female = factor(female),
    parental_divorce = factor(parental_divorce),
    hazard = 1 / (1 + exp(-.fitted)),
    survival = cumprod(1 - hazard),
    .by = c(female, parental_divorce)
  ) |>
  pivot_longer(cols = c(hazard, survival), names_to = "estimate") |>
  ggplot(aes(x = period, y = value, linetype = female, colour = parental_divorce)) +
    geom_line() +
    facet_wrap(vars(estimate), ncol = 1, scales = "free_y") +
    scale_x_continuous(breaks = seq(0, 40, by = 5), limits = c(0, 40))  +
    ggh4x::facetted_pos_scales(
      y = list(
        estimate == "hazard" ~ scale_y_continuous(limits = c(0, .04)),
        estimate == "survival" ~ scale_y_continuous(limits = c(0, 1))
      )
    )
```

## 12.4 The Linear Additivity Assumption: Uncovering Violations and Simple Solutions

Figure 12.6, page 445:

```{r}
# Raw
first_arrest |>
  group_by(period, abused, black) |>
  summarise(
    total = n(),
    event = sum(event),
    proportion = event / total,
    proportion = if_else(proportion == 0, NA, proportion),
    logit = log(proportion / (1 - proportion))
  ) |>
  ungroup() |>
  mutate(across(c(abused, black), factor)) |>
  na.omit() |>
  ggplot(aes(x = period, y = logit, colour = abused, group = abused)) +
    geom_line() +
    scale_x_continuous(breaks = 7:19, limits = c(7, 19)) +
    scale_y_continuous(limits = c(-7, -2)) +
    facet_wrap(vars(black), labeller = label_both)

# Model
first_arrest_fit <- glm(
  event ~ factor(period) + abused + black + abused:black,
  family = binomial(link = "logit"),
  data = first_arrest
)

first_arrest_fit |>
  augment(
    newdata = expand_grid(period = 8:18, abused = c(0, 1), black = c(0, 1))
  ) |>
  ggplot(
    aes(
      x = period, y = .fitted, colour = factor(abused), linetype = factor(black)
    )
  ) +
  geom_line() +
  scale_x_continuous(breaks = 7:19, limits = c(7, 19)) +
  scale_y_continuous(limits = c(-8, -2))
```

Table 12.4, page 449:

```{r}
model_A <- update(first_depression_fit_2, . ~ . + siblings)
model_B <- update(
  first_depression_fit_2, 
  . ~ . +
    between(siblings, 1, 2) +
    between(siblings, 3, 4) +
    between(siblings, 5, 6) +
    between(siblings, 7, 8) +
    between(siblings, 9, Inf)
)
model_C <- update(first_depression_fit_2, . ~ . + bigfamily)

tidy(model_A)
tidy(model_B)
tidy(model_C)
```

## 12.5 The proportionality assumption: Uncovering violations and simple solutions

Figure 12.8, page 458:

```{r}
# Raw
math_dropout |>
  group_by(term, woman) |>
  summarise(
    total = n(),
    event = sum(event),
    proportion = event / total,
    proportion = if_else(proportion == 0, NA, proportion),
    logit = log(proportion / (1 - proportion))
  ) |>
  ungroup() |>
  mutate(across(c(woman), factor)) |>
  na.omit() |>
  ggplot(aes(x = term, y = logit, colour = woman)) +
    geom_line()

# Models
model_A <- glm(
  event ~ -1 + factor(term) + woman,
  family = binomial(link = "logit"),
  data = math_dropout
)
model_B <- glm(
  event ~ -1 + factor(term) + factor(term):woman,
  family = binomial(link = "logit"),
  data = math_dropout
)
model_C <- update(model_A, . ~ . + woman:I(term - 1))

map_df(
  list(model_A = model_A, model_B = model_B, model_C = model_C),
  \(.x) {
    .x |>
      augment(newdata = expand_grid(term = 1:5, woman = c(0, 1))) |>
      mutate(hazard = 1 / (1 + exp(-.fitted)))
  },
  .id = "model"
) |>
  ggplot(aes(x = term, y = hazard, colour = factor(woman))) +
    geom_line() +
    facet_wrap(vars(model))
```

Table 12.5, page 459:

```{r}
tidy(model_A)
tidy(model_B)
tidy(model_C)
```

## 12.6 The No Unobserved Heterogeneity Assumption: No Simple Solution

## 12.7 Residual Analysis

Table 12.6, page 465:

```{r}
first_sex_fit <- glm(
  event ~ -1 + factor(grade) + parental_transition + parental_antisociality,
  family = binomial(link = "logit"),
  data = first_sex_pp
)

first_sex_fit |>
  augment(data = first_sex_pp, type.residuals = "deviance") |>
  select(id:parental_antisociality, .resid) |>
  filter(id %in% c(22, 112, 166, 89, 102, 87, 67, 212)) |>
  pivot_wider(
    id_cols = id,
    names_from = grade,
    names_prefix = "grade_",
    values_from = .resid
  )
```

Figure 12.8, page 467:

```{r}
first_sex_fit |>
  augment(data = first_sex_pp, type.residuals = "deviance") |>
  ggplot(aes(x = id, y = .resid)) +
    geom_point() +
    geom_hline(yintercept = 0)

first_sex_fit |>
  augment(data = first_sex_pp, type.residuals = "deviance") |>
  group_by(id) |>
  summarise(ss.deviance = sum(.resid^2)) |>
  ggplot(aes(x = id, y = ss.deviance)) +
    geom_point()
```
