---
title: "Chapter 11: Fitting basic discrete-time hazard models"
---

::: {.alert .alert-warning}
This chapter is under construction.
:::

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(
  panel.grid.major = ggplot2::element_blank(),
  panel.grid.minor = ggplot2::element_blank()
)
```

```{r setup}
library(alda)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(survival)
library(broom)
```

## 11.1 Toward a Statistical Model for DiscreteTime Hazard

Several examples in this chapter rely on the following:

```{r}
first_sex_fit <- survfit(Surv(grade, 1 - censor) ~ 1, data = first_sex)

first_sex_pt <- c(0, 1) |>
  map_dfr(
    \(.x) {
      first_sex_fit_subset <- update(
        first_sex_fit, subset = (parental_transition == .x)
      )
      
      first_sex_fit_subset |>
        survfit0(start.time = 6) |>
        tidy() |>
        rename(survival_probability = estimate) |>
        mutate(
          hazard_probability = n.event / n.risk,
          odds = hazard_probability / (1 - hazard_probability),
          log_odds = log(odds)
        ) |>
        select(-starts_with("conf"), -std.error) |>
        rename(grade = time) |>
        pivot_longer(
          cols = c(survival_probability, hazard_probability, odds, log_odds),
          values_to = "estimate"
        ) |>
        # The figure doesn't include data for grade 6 in the hazard function.
        filter(
          !(name %in% c("hazard_probability", "odds", "log_odds") & grade == 6)
        )
    },
    .id = "parental_transition"
  )
```

Figure 11.1, page 359:

```{r}
first_sex_pt |>
  filter(name %in% c("survival_probability", "hazard_probability")) |>
  ggplot(aes(x = grade, y = estimate, colour = parental_transition)) +
    geom_hline(
      aes(yintercept = .5),
      data = tibble(name = "survival_probability"),
      alpha = .25,
      linetype = 2
    ) +
    geom_line() +
    scale_x_continuous(breaks = 6:12) +
    coord_cartesian(xlim = c(6, 12)) +
    facet_wrap(vars(name), ncol = 1, scales = "free_y") +
    ggh4x::facetted_pos_scales(
      y = list(
        name == "hazard_probability" ~ scale_y_continuous(limits = c(0, .5)),
        name == "survival_probability" ~
          scale_y_continuous(breaks = c(0, .5, 1), limits = c(0, 1))
      )
    )
```

Table 11.1, page 360:

```{r}
# First two sections of the table
first_sex_pt |>
  filter(grade != 6, !(name %in% c("odds", "log_odds"))) |>
  pivot_wider(names_from = name, values_from = estimate) |>
  select(everything(), -n.censor, hazard_probability, survival_probability)

# Last section
first_sex_fit |>
  tidy() |>
  rename(survival_probability = estimate) |>
  mutate(
    hazard_probability = n.event / n.risk,
    .before = survival_probability
  ) |>
  select(-starts_with("conf"), -std.error, -n.censor) |>
  rename(grade = time)
```

Figure 11.2, page 363:

```{r}
first_sex_pt |>
  filter(name %in% c("hazard_probability", "odds", "log_odds")) |>
  mutate(
    name = factor(name, levels = c("hazard_probability", "odds", "log_odds"))
  ) |>
  ggplot(aes(x = grade, y = estimate, colour = parental_transition)) +
    geom_line() +
    scale_x_continuous(breaks = 6:12) +
    coord_cartesian(xlim = c(6, 12)) +
    facet_wrap(vars(name), ncol = 1, scales = "free_y") +
    ggh4x::facetted_pos_scales(
      y = list(
        name %in% c("hazard_probability", "odds") ~
          scale_y_continuous(limits = c(0, 1)),
        name == "log_odds" ~ scale_y_continuous(limits = c(-4, 0))
      )
    )
```

Figure 11.3, page 366:

```{r}
# Transform to person-period format.
first_sex_pp <- first_sex |>
  rename(grades = grade) |>
  reframe(
    grade = 7:max(grades),
    event = if_else(grade == grades & censor == 0, 1, 0),
    parental_transition,
    parental_antisociality,
    .by = id
  )

# Fit models for each panel.
first_sex_fit_11.3a <- glm(
  event ~ parental_transition, family = "binomial", data = first_sex_pp
)
first_sex_fit_11.3b <- update(first_sex_fit_11.3a, . ~ . + grade)
first_sex_fit_11.3c <- update(first_sex_fit_11.3a, . ~ . + factor(grade))

# Plot:
map_df(
  list(a = first_sex_fit_11.3a, b = first_sex_fit_11.3b, c = first_sex_fit_11.3c),
  \(.x) augment(.x, newdata = first_sex_pp),
  .id = "model"
) |>
  ggplot(aes(x = grade, y = .fitted, colour = factor(parental_transition))) +
  geom_line() +
  geom_point(
    aes(y = estimate),
    data = first_sex_pt |>
      mutate(parental_transition = as.numeric(parental_transition) - 1) |>
      filter(name == "log_odds")
  ) +
  coord_cartesian(ylim = c(-4, 0)) +
  facet_wrap(vars(model), ncol = 1, labeller = label_both) +
  labs(
    y = "logit(hazard)",
    colour = "parental_transition"
  )
```

## 11.2 A Formal Representation of the Population Discrete-Time Hazard Model

Figure 11.4, page 374:

```{r}
# Panel A:
first_sex_fit_11.3c |>
  augment(newdata = first_sex_pp) |>
  ggplot(aes(x = grade, y = .fitted, colour = factor(parental_transition))) +
  geom_line() +
  coord_cartesian(ylim = c(-4, 0))

# Panel B:
first_sex_fit_11.4b <- update(
  first_sex_fit_11.3c,
  . ~ . + parental_transition * factor(grade)
)

first_sex_fit_11.4b |>
  augment(newdata = first_sex_pp) |>
  ggplot(aes(x = grade, y = exp(.fitted), colour = factor(parental_transition))) +
  geom_line() +
  coord_cartesian(ylim = c(0, 1))

# Panel C:
first_sex_fit_11.4b |>
  augment(newdata = first_sex_pp, type.predict = "response") |>
  ggplot(aes(x = grade, y = .fitted, colour = factor(parental_transition))) +
  geom_line()
```

## 11.3 Fitting a Discrete-Time Hazard Model to Data

Figure 11.5:

```{r}

```

Table 11.3, page 386:

```{r}
model_A <- glm(
  event ~ factor(grade) - 1, family = "binomial", data = first_sex_pp
)
model_B <- update(model_A, . ~ . + parental_transition)
model_C <- update(model_A, . ~ . + parental_antisociality)
model_D <- update(model_B, . ~ . + parental_antisociality)

anova(model_B)
anova(model_C)
# Deviance tests are sequential so the order of terms matters. To test
# parental_transition and parental_antisociality, the model needs to be
# fit twice, once with each as the last term.
anova(update(model_C, . ~ . + parental_transition))
anova(model_D)
```

## 11.4 Interpreting Parameter Estimates

Table 11.4, page 388:

```{r}
model_A |>
  tidy() |>
  select(term, estimate) |>
  mutate(
    odds = exp(estimate),
    hazard = 1 / (1 + exp(-estimate))
  )
```

## 11.5 Displaying Fitted Hazard and Survivor Functions

Table 11.5, page 392:

```{r}
model_B_tidy <- model_B |>
  augment(
    newdata = expand_grid(grade = 7:12, parental_transition = 0:1)
  ) |>
  mutate(
    hazard = 1 / (1 + exp(-.fitted)),
    survival = cumprod(1 - hazard),
    .by = parental_transition
  )

model_B_tidy
```

Figure 11.6, page 393:

```{r}
# FIXME: should use survfit0() for the survival panel so time starts at 6.
model_B_tidy |>
  pivot_longer(cols = .fitted:survival) |>
  ggplot(aes(x = grade, y = value, colour = factor(parental_transition))) +
  geom_line() +
  facet_wrap(vars(name), ncol = 1, scales = "free_y")
```

Figure 11.7, page 395:

```{r}
prototypical_males <- tibble(
  id = rep(1:6, times = length(7:12)),
  expand_grid(
    grade = 7:12,
    parental_transition = c(0, 1),
    parental_antisociality = -1:1
  )
)

prototypical_first_sex <- tibble(
  log_odds = predict(
    model_D,
    prototypical_males
  ),
  hazard = 1 / (1 + exp(-log_odds))
)

grade_six <- tibble(
  id = 1:6,
  grade = 6,
  expand_grid(
    parental_transition = c(0, 1),
    parental_antisociality = -1:1
  ),
  log_odds = NA,
  hazard = NA,
  survival = 1
)

prototypical_males |>
  bind_cols(prototypical_first_sex) |>
  mutate(survival = cumprod(1 - hazard), .by = id) |>
  add_row(grade_six) |>
  pivot_longer(cols = c(hazard, survival)) |>
  ggplot(aes(x = grade, y = value, group = id)) +
    geom_line(
      aes(
        colour = factor(parental_antisociality),
        linetype = factor(parental_transition)
      )
    ) +
    scale_colour_grey(start = 0, end = 0.75) +
    facet_wrap(
      vars(name),
      ncol = 1,
      scales = "free_y"
    )
```
