---
title: "Chapter 10: Describing discrete-time event occurrence data"
---

::: {.alert .alert-warning}
This chapter is under construction.
:::

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(
  panel.grid.major = ggplot2::element_blank(),
  panel.grid.minor = ggplot2::element_blank()
)
```

```{r setup}
library(alda)
library(dplyr)
library(tidyr)
library(ggplot2)
library(survival)
library(broom)
```

## 10.1 The Life Table

Table 10.1, page 327:

```{r}
# A life table is the tabular form of a survival curve, so begin by fitting a
# Kaplan-Meir curve to the data.
teachers_fit <- survfit(Surv(years, 1 - censor) ~ 1, data = teachers)

table_10.1 <- teachers_fit |>
  # Add a starting time (time 0) for the table.
  survfit0() |>
  tidy() |>
  # The summary of the fit gives most of what we want, but to match Table 10.1
  # we need to do a little more wrangling.
  select(-c(std.error:conf.low)) |>
  mutate(
    interval = paste0("[", time, ", ", time + 1, ")"),
    haz.estimate = n.event / n.risk
  ) |>
  rename(year = time, surv.estimate = estimate) |>
  relocate(
    year, interval, n.risk, n.event, n.censor, haz.estimate, surv.estimate
  )

table_10.1
```

## 10.2 A Framework for Characterizing the Distribution of Discrete-Time Event Occurrence Data

Figure 10.1, page 333:

```{r}
ggplot(table_10.1, aes(x = year, y = haz.estimate)) +
  geom_line() +
  scale_x_continuous(breaks = 0:13, limits = c(1, 13)) +
  scale_y_continuous(breaks = c(0, .05, .1, .15), limits = c(0, .15)) +
  coord_cartesian(xlim = c(0, 13))

# First interpolate median lifetime
median_lifetime <- table_10.1 |>
  # Get the row indices for the first survival estimate immediately below and
  # immediately above 0.5. This will only work correctly if the values are in
  # descending order, otherwise min() and max() must be swapped. By default, the
  # survival estimates are in descending order, however, I've added the
  # redundant step of ensuring they are here for demonstration purposes.
  arrange(desc(surv.estimate)) |>
  slice(min(which(surv.estimate <= .5)), max(which(surv.estimate >= .5))) |>
  select(year, surv.estimate) |>
  # Linearly interpolate between the two values of the survival estimates that
  # bracket .5 following Miller's (1981) equation.
  summarise(
    year =
      min(year) +
      ((max(surv.estimate) - .5) /
       (max(surv.estimate) - min(surv.estimate))) *
      ((min(year) + 1) - min(year)),
    surv.estimate = .5
  )

ggplot(table_10.1, aes(x = year, y = surv.estimate)) +
  geom_line() +
  geom_segment(
    aes(xend = year, y = 0, yend = .5), data = median_lifetime, linetype = 2
  ) +
  geom_segment(
    aes(xend = 0, yend = .5), data = median_lifetime, linetype = 2
  ) +
  scale_x_continuous(breaks = 0:13) +
  scale_y_continuous(breaks = c(0, .5, 1), limits = c(0, 1)) +
  coord_cartesian(xlim = c(0, 13))
```

## 10.3 Developing Intuition About Hazard Functions, Survivor Functions, and Median Lifetimes

Figure 10.2, page 340:

```{r}
relapse_fit <- survfit(Surv(weeks, 1 - censor) ~ 1, data = cocaine_relapse_1)
relapse_tidy <- tidy(relapse_fit)
relapse_summary <- glance(relapse_fit)
```

## 10.4 Quantifying the Effects of Sampling Variation

Table 10.2, page 349:

```{r}
summary(teachers_fit)

teachers_fit |>
  tidy() |>
  mutate(
    # The tidy() method for survfit objects returns the standard error for the
    # cumulative hazard instead of the survival probability. Multiplying the
    # survival estimate with the cumulative hazard's standard error will return
    # the standard error for the survival probability. Note that it is unlikely
    # the tidy() method will ever change to return the the standard error for
    # the survival probability instead. See:
    # - https://github.com/tidymodels/broom/pull/1162
    # Other transformations of the survival probability can be found here:
    # - https://stat.ethz.ch/pipermail/r-help/2014-June/376247.html
    surv.std.error = estimate * std.error,
    haz.estimate = n.event / n.risk,
    haz.std.error = sqrt(haz.estimate * (1 - haz.estimate) / n.risk),
    sqrt = (std.error)^2 / (estimate)^2
  ) |>
  select(
    year = time,
    n.risk,
    haz.estimate,
    haz.std.error,
    surv.estimate = estimate,
    sqrt,
    surv.std.error
  )
```

## 10.5 A Simple and Useful Strategy for Constructing the Life Table

Figure 10.4, page 353:

```{r}
filter(teachers, id %in% c(20, 126, 129))

teachers_pp <- teachers |>
  reframe(
    year = 1:max(years),
    event = if_else(year == years & censor == 0, 1, 0),
    .by = id
  )

teachers_pp |>
  filter(id %in% c(20, 126, 129)) |>
  print(n = 27)
```

Table 10.3, page 355:

```{r}
teachers_pp |>
  group_by(year) |>
  count(event) |>
  pivot_wider(names_from = event, names_prefix = "event_", values_from = n) |>
  mutate(
    total = event_0 + event_1,
    p.event_1 = event_1 / total
  )
```
