---
title: "Chapter 8: Modeling change using covariance structure analysis"
---

::: {.alert .alert-warning}
This chapter is under construction.
:::

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(alda)
library(dplyr)
library(tidyr)
library(lavaan)
library(tidySEM)
```

Note: Covariance Structure Analysis is just another name for structural equation modelling (SEM).

## 8.1 The General Covariance Structure Model

## 8.2 The Basics of Latent Growth Modeling

Table 8.1, page 282:

```{r}
alcohol_use_2_wide <- alcohol_use_2 |>
  pivot_wider(names_from = time, values_from = c(alcohol_use, peer_pressure))

alcohol_use_2_wide

# Means
alcohol_use_2_wide |>
  summarise(across(female:peer_pressure_2, mean)) |>
  glimpse()

# Covariances
cov(select(alcohol_use_2_wide, -c(id, female)))
```

Table 8.2, page 289, Model A:

```{r}
# Model A: Unconditional model
model_A <- ("
  # Intercept and slope with fixed coefficients
  intercept =~ 1*alcohol_use_0 + 1*alcohol_use_1 + 1*alcohol_use_2
  slope     =~ 0*alcohol_use_0 + .75*alcohol_use_1 + 1.75*alcohol_use_2
")

model_A_fit <- growth(
  model_A, data = alcohol_use_2_wide, estimator = "ml", mimic = "Mplus"
)

summary(model_A_fit)
fitMeasures(model_A_fit, c("chisq", "df", "pvalue", "cfi", "rmsea"))
```

Figure 8.2, Model A:

```{r}
lay <- get_layout(
  NA, "intercept", NA,  "slope", NA,
  "alcohol_use_0", NA,  "alcohol_use_1", NA, "alcohol_use_2",
  rows = 2
)

graph_sem(model_A_fit, layout = lay)
```

Table 8.2, page 289, Model B:

```{r}
# Model B: Adding female as a time-invariant predictor
model_B <- ("
  # Intercept and slope with fixed coefficients
  intercept =~ 1*alcohol_use_0 + 1*alcohol_use_1 + 1*alcohol_use_2
  slope =~ 0*alcohol_use_0 + .75*alcohol_use_1 + 1.75*alcohol_use_2
  
  # Regressions
  intercept ~ female
  slope ~ female
")

model_B_fit <- growth(
  model_B, data = alcohol_use_2_wide, estimator = "ml", mimic = "Mplus"
)

summary(model_B_fit)
fitMeasures(model_B_fit, c("chisq", "df", "pvalue", "cfi", "rmsea"))
```

Comparison with baseline model for Model B:

```{r}
# Baseline for Model B (not shown in table)
model_B_baseline <- ("
  # Intercept and slope with fixed coefficients
  intercept =~ 1*alcohol_use_0 + 1*alcohol_use_1 + 1*alcohol_use_2
  slope =~ 0*alcohol_use_0 + .75*alcohol_use_1 + 1.75*alcohol_use_2
  
  # Regressions
  intercept ~ 0*female
  slope ~ 0*female
  
  alcohol_use_0 ~ 0*1
  alcohol_use_1 ~ 0*1
  alcohol_use_2 ~ 0*1
")

model_B_baseline_fit <- growth(
  model_B_baseline, data = alcohol_use_2_wide, estimator = "ml", mimic = "Mplus"
)

anova(model_B_baseline_fit, model_B_fit)
```

Figure 8.2, Model B:

```{r}
lay <- get_layout(
  NA, NA, "female", NA, NA,
  NA, "intercept", NA,  "slope", NA,
  "alcohol_use_0", NA,  "alcohol_use_1", NA, "alcohol_use_2",
  rows = 3
)

graph_sem(model_B_fit, layout = lay)
```

Table 8.2, page 289, Model C:

```{r}
# Model C: Model B but with slope fixed to zero
model_C <- ("
  # Intercept and slope with fixed coefficients
  intercept =~ 1*alcohol_use_0 + 1*alcohol_use_1 + 1*alcohol_use_2
  slope =~ 0*alcohol_use_0 + .75*alcohol_use_1 + 1.75*alcohol_use_2
  
  # Regressions
  intercept ~ female
  slope ~ 0*female
")

model_C_fit <- growth(
  model_C, data = alcohol_use_2_wide, estimator = "ml", mimic = "Mplus"
)

summary(model_C_fit)
fitMeasures(model_C_fit, c("chisq", "df", "pvalue", "cfi", "rmsea"))
```

Figure 8.2, Model C (same as Model B but with `slope ~ 0*female`):

```{r}
graph_sem(model_C_fit, layout = lay)
```

Table 8.2, page 289, Model D:

```{r}
# Model D: Adding peer_pressure as a time-varying predictor
model_D <- ("
  # Intercept and slope with fixed coefficients
  alc_intercept =~ 1*alcohol_use_0 + 1*alcohol_use_1 + 1*alcohol_use_2
  alc_slope =~ 0*alcohol_use_0 + .75*alcohol_use_1 + 1.75*alcohol_use_2
  
  peer_intercept =~ 1*peer_pressure_0 + 1*peer_pressure_1 + 1*peer_pressure_2
  peer_slope =~ 0*peer_pressure_0 + .75*peer_pressure_1 + 1.75*peer_pressure_2
  
  # Regressions
  alc_intercept ~ start(.8)*peer_intercept + start(.08)*peer_slope
  alc_slope ~ start(-.1)*peer_intercept + start(.6)*peer_slope
  
  # Time-varying covariances
  alcohol_use_0 ~~ peer_pressure_0
  alcohol_use_1 ~~ peer_pressure_1
  alcohol_use_2 ~~ peer_pressure_2
  
  # Fix intercepts to zero
  alcohol_use_0 ~ 0*1
  alcohol_use_1 ~ 0*1
  alcohol_use_2 ~ 0*1
  peer_pressure_0 ~ 0*1
  peer_pressure_1 ~ 0*1
  peer_pressure_2 ~ 0*1

")

model_D_fit <- growth(
  model_D, data = alcohol_use_2_wide, estimator = "ml", mimic = "Mplus"
)

summary(model_D_fit)
fitMeasures(model_D_fit, c("chisq", "df", "pvalue", "cfi", "rmsea"))
```

Comparison with baseline model for Model D:

```{r}
# Baseline for Model D (not shown in table)
model_D_baseline <- ("
  # Intercepts and slopes with fixed coefficients
  alc_intercept =~ 1*alcohol_use_0 + 1*alcohol_use_1 + 1*alcohol_use_2
  alc_slope =~ 0*alcohol_use_0 + .75*alcohol_use_1 + 1.75*alcohol_use_2
  
  peer_intercept =~ 1*peer_pressure_0 + 1*peer_pressure_1 + 1*peer_pressure_2
  peer_slope =~ 0*peer_pressure_0 + .75*peer_pressure_1 + 1.75*peer_pressure_2
  
  # Regressions
  alc_intercept ~ 0*peer_intercept + 0*peer_slope
  alc_slope ~ 0*peer_intercept + 0*peer_slope
  
  # Time-varying covariances
  alcohol_use_0 ~~ peer_pressure_0
  alcohol_use_1 ~~ peer_pressure_1
  alcohol_use_2 ~~ peer_pressure_2
  
  alcohol_use_0 ~ 0*1
  alcohol_use_1 ~ 0*1
  alcohol_use_2 ~ 0*1
  peer_pressure_0 ~ 0*1
  peer_pressure_1 ~ 0*1
  peer_pressure_2 ~ 0*1
")

model_D_baseline_fit <- growth(
  model_D_baseline, data = alcohol_use_2_wide, estimator = "ml", mimic = "Mplus"
)

anova(model_D_baseline_fit, model_D_fit)
```

Figure 8.2, Model D:

```{r}
lay <- get_layout(
  "peer_pressure_0", NA,  "peer_pressure_1", NA, "peer_pressure_2",
  NA, "peer_intercept", NA,  "peer_slope", NA,
  NA, "alc_intercept", NA,  "alc_slope", NA,
  "alcohol_use_0", NA,  "alcohol_use_1", NA, "alcohol_use_2",
  rows = 4
)

graph_sem(model_D_fit, layout = lay)
```

## 8.3 Cross-Domain Analysis of Change

## 8.4 Extensions of Latent Growth Modeling
